\vspace{-0.2in} Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-05 main.tex
Applied Hossein's corrections

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the $d$ pixels of the training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$. 
%Since $A$ contains many images per user taken under different illuminations, 
%$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

In Lagrange Multiplier Theory \cite{BertsekasD2003}, if there exists a Lagrangian $y^*$ that
satisfies the second-order sufficiency conditions for optimality, then for a sufficiently large $\mu$, the optimal
$\ell_1$-min solution also minimizes
\begin{equation}
\xx^* = \arg \min L_\mu(\xx,\yy^*).
\label{eq:optimal-ALM}
\end{equation}

@@

Note that in the iterative procedure \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrangian $L_{\mu_k} (\xx,\yy_k)$ compared to solving the original problem
\eqref{eq:l1min} directly. In fact, this problem can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the generic ALM $\ell_1$-min algorithm. \footnote{For conciseness, we
only present the ALM algorithm in the primal domain. There also
exist implementations in the dual domain \cite{YangJ2009,YangA2010-ICIP}.}

 \begin{algorithm}[h]
\caption{Augmented Lagrangian Method (ALM)}
\small
{\bf INPUT:} $\bb \in \Re^m$, $A=[A_1,\cdots, A_K] \in \Re^{m \times n}$, $\tau\leftarrow \max\mbox{eig}(A^TA)$, and constant $\rho>1$.
\begin{algorithmic}[1]
\WHILE{not converged ($k = 1,2,\ldots$)} 
\STATE $t_1 \leftarrow 1$, $\zz_1 \leftarrow \xx_k$, $\uu_1 \leftarrow \xx_k$ 
\WHILE{not converged ($l = 1,2,\ldots$)} 
\STATE $\uu_{l+1}  \leftarrow \mbox{shrink}(\zz_l - \frac{1}{\tau}A^T(A\zz_l - \bb - \frac{1}{\mu_k}\yy_k), \frac{1}{\mu_k\tau})$
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 + \sqrt{1+4t_l^2})$
\STATE $\zz_{l+1} \leftarrow \uu_{l+1}+ \frac{t_l - 1}{t_{l+1}}(\uu_{l+1} - \uu_l)$ 
\ENDWHILE 
\STATE $\xx_{k+1} \leftarrow \uu_{l+1}$ 
\STATE $\yy_{k+1} \leftarrow \yy_k + \mu_k (\bb - A\xx_{k+1})$ 
\STATE $\mu_{k+1} \leftarrow \rho\cdot\mu_k$ 
\ENDWHILE 
\end{algorithmic}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is inefficient since it is a non-convex problem
and may exhibit local minima.
However, given a good initial estimate of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

In both implementations, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency (for the first implementation),
but are also available in single-threaded versions (for the second implementation). For 
operations that are not optimized by Intel in-house libraries, such as the shrinkage
operator in step 7 of Algorithm \ref{alg:iterative_alignment}, we achieve 
vector-level concurrency via
the automatic vectorization facilities of the Intel C++ compiler (ICC) \cite{dulong1999overview}.
To achieve core-level concurrency we make use of the Open MP API. \cite{dagum2002openmp} 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

@@

\begin{algorithm}[t]
\caption{\bf (Face Recognition via ALM)} \label{alg:alm_rec} 
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\y_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}(\bb - A\x_k +
\frac{1}{\mu}\y_k, \frac{1}{\mu})$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}(\z_l +
\frac{1}{\gamma}A^T(\bb - A\z_l - \e_{k+1} +
\frac{1}{\mu}\y_k), \frac{1}{\mu\gamma})$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 +
\sqrt{1+4t_l^2})$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\y_{k+1} \leftarrow \y_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\end{algorithm}

@@

The first experiment compares the performance of our proposed CPU and GPU
implementations of the general $\ell_1$-min solver (Algorithm \eqref{alg:alm}).
The $d \times n$ measurement matrix $A$ is a random Gaussian matrix, with each
entry generated from the standard normal distribution and normalized to unit
column norm.  The ratio of $d/n$ is fixed at $1/2$ with $n$ varying from 1000 to 8000.
The ground truth signal, $\xx_0$ has a sparsity rate of 10\% of d with
elements sampled from a normal distribution and normalized to unit column
norm.  Because the ground truth signal $\xx_0$ is known, the algorithm
terminates when $\|\bx-\bx_0\| < \tau$ with $\tau=10^{-3}$.  The measurement
vector is generated by $\bb = A \xx_0$.

>>>>

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the $m$ pixels of the $n_i$ training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{m\times n_1}, \cdots, A_K\in\Re^{m\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{m\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^m$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$. 
%Since $A$ contains many images per user taken under different illuminations, 
%$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

In Lagrange Multiplier Theory \cite{BertsekasD2003}, if there exists a multiplier $\y^*$ that
satisfies the second-order sufficiency conditions for optimality, then for a sufficiently large $\mu$, the optimal
$\ell_1$-min solution also minimizes
\begin{equation}
\xx^* = \arg \min L_\mu(\xx,\yy^*).
\label{eq:optimal-ALM}
\end{equation}

@@

Note that in the iterative procedure \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrangian $L_{\mu_k} (\xx,\yy_k)$ compared to solving the original problem
\eqref{eq:l1min} directly. In fact, this problem can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the generic ALM $\ell_1$-min algorithm. \footnote{For conciseness, we
only present the ALM algorithm in the primal domain. There also
exist implementations in the dual domain \cite{YangJ2009,YangA2010-ICIP}.}
The soft-thresholding algorithm involves the element-wise operator $\shrink(x) = \sign(x)\max(|x|,0)$.

 \begin{algorithm}[h]
\caption{Augmented Lagrangian Method (ALM)}
\small
{\bf INPUT:} $\bb \in \Re^m$, $A=[A_1,\cdots, A_K] \in \Re^{m \times n}$, $\tau\leftarrow \max\mbox{eig}(A^TA)$, and constant $\rho>1$.
\begin{algorithmic}[1]
\WHILE{not converged ($k = 1,2,\ldots$)} 
\STATE $t_1 \leftarrow 1$, $\zz_1 \leftarrow \xx_k$, $\uu_1 \leftarrow \xx_k$ 
\WHILE{not converged ($l = 1,2,\ldots$)} 
\STATE $\uu_{l+1}  \leftarrow \shrink(\zz_l - \frac{1}{\tau}A^T(A\zz_l - \bb - \frac{1}{\mu_k}\yy_k), \frac{1}{\mu_k\tau})$
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 + \sqrt{1+4t_l^2})$
\STATE $\zz_{l+1} \leftarrow \uu_{l+1}+ \frac{t_l - 1}{t_{l+1}}(\uu_{l+1} - \uu_l)$ 
\ENDWHILE 
\STATE $\xx_{k+1} \leftarrow \uu_{l+1}$ 
\STATE $\yy_{k+1} \leftarrow \yy_k + \mu_k (\bb - A\xx_{k+1})$ 
\STATE $\mu_{k+1} \leftarrow \rho\cdot\mu_k$ 
\ENDWHILE 
\end{algorithmic}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is inefficient since it is a non-convex problem
and may exhibit local minima.
However, given a good initial estimate of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \shrink\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

In both implementations, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency (for the first implementation),
but are also available in single-threaded versions (for the second implementation). For 
operations that are not optimized by Intel in-house libraries, such as the soft thresholding
operator in step 7 of Algorithm \ref{alg:iterative_alignment}, we achieve 
vector-level concurrency via
the automatic vectorization facilities of the Intel C++ compiler (ICC) \cite{dulong1999overview}.
To achieve core-level concurrency we make use of the Open MP API. \cite{dagum2002openmp} 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

@@

\begin{algorithm}[t]
\caption{\bf (Face Recognition via ALM)} \label{alg:alm_rec} 
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\y_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \shrink(\bb - A\x_k +
\frac{1}{\mu}\y_k, \frac{1}{\mu})$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \shrink(\z_l +
\frac{1}{\gamma}A^T(\bb - A\z_l - \e_{k+1} +
\frac{1}{\mu}\y_k), \frac{1}{\mu\gamma})$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 +
\sqrt{1+4t_l^2})$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\y_{k+1} \leftarrow \y_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\end{algorithm}

@@

The first experiment compares the performance of our proposed CPU and GPU
implementations of the general $\ell_1$-min solver (Algorithm \eqref{alg:alm}).
The $m \times n$ measurement matrix $A$ is a random Gaussian matrix, with each
entry generated from the standard normal distribution and normalized to unit
column norm.  The ratio of $m/n$ is fixed at $1/2$ with $n$ varying from 1000 to 8000.
The ground truth signal, $\xx_0$ has a sparsity rate of 10\% of $m$ with
elements sampled from a normal distribution and normalized to unit column
norm.  Because the ground truth signal $\xx_0$ is known, the algorithm
terminates when $\|\bx-\bx_0\| < \tau$ with $\tau=10^{-3}$.  The measurement
vector is generated by $\bb = A \xx_0$.
